% Configuring document class and essential packages
\documentclass[a4paper,12pt]{article}

% Including standard LaTeX packages for formatting and mathematics
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{mathtools}
\usepackage{geometry}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{parskip}
\usepackage{tocloft}
\usepackage{tikz} % For simple diagrams if needed

% Defining custom commands for mathematical notation
\newcommand{\C}{\mathcal{C}}
\newcommand{\M}{\mathcal{M}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\Ocal}{\mathcal{O}}
\newcommand{\U}{\mathcal{U}}
\newcommand{\V}{\mathcal{V}}
\newcommand{\p}{\mathbf{p}}
\newcommand{\lvec}{\mathbf{l}}
\newcommand{\E}{\mathrm{E}}
\newcommand{\Scal}{\mathcal{S}}
\newcommand{\vvec}{\vec{v}}
\DeclareMathOperator{\hocolim}{hocolim}

% Setting page geometry
\geometry{left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm}

% Beginning the document
\begin{document}

% Title and author
\title{Geometric Memory as Trajectory Influence on a Generative Substrate}
\author{Flyxion}
\date{September 18, 2025}
\maketitle

% Abstract
\begin{abstract}
This essay formalizes memory as trajectories influencing a generative substrate, modeled as a smooth manifold of meaning-states. Trajectories bias future dynamics via a nonlocal-in-time flow, integrating past influences through kernel-weighted parallel transport. The substrate evolves under a Riemannian metric, with curvature and torsion shaping convergence and reinterpretation. Plasticity updates the metric and connection, embedding habitual paths as low-resistance channels. The model integrates with the Relativistic Scalar Vector Plenum (RSVP) framework, mapping scalar density ($\Phi$), vector flows ($\vvec$), and entropy ($\Scal$) to memory processes. Semantic ledging and virtual memory extend the model, handling mismatched cues and dynamic scaling, inspired by geometric memory management \citep{kuijper2021geometric}. Testable predictions link curvature to habits, kernel perturbations to recall effects, and torsion to context-dependent reinterpretation, with applications to neuroscience (fMRI/EEG), large language models (LLMs), and semantic infrastructures. A discrete implementation aligns with transformer architectures, offering a computationally tractable framework for experimentation. Empirical validation through behavioral tasks and AI ablations demonstrates the model's predictive power, while limitations in quantum regimes and qualia highlight avenues for future extension.
\end{abstract}

% Table of Contents
\tableofcontents

\section{Introduction}
Traditional memory models, often likened to a ``file cabinet'' where static records are stored and retrieved, fail to capture the dynamic, reconstructive nature of memory in biological and computational systems \citep{mcclelland1995why}. Empirical findings from cognitive psychology and neuroscience demonstrate that recall is not a simple replay of fixed traces but a generative act, where fragments of prior experience are recombined and filtered through ongoing context and current goals. For instance, the serial position effect—where items at the beginning (primacy) and end (recency) of a list are better recalled—suggests that memory is influenced by both long-term consolidation and short-term rehearsal, rather than uniform storage \citep{murdock1962serial}. This essay proposes a geometric model in which memory is understood as a trajectory on a generative substrate—a smooth manifold of meaning-states—whose past paths bias future dynamics through nonlocal influence mechanisms.

In this model, memories are not fetched from a static archive but are regenerated by kernel-weighted integrals that transport prior trajectory increments into the present state. The manifold’s geometry provides the structural constraints: the Riemannian metric encodes semantic distance, curvature shapes convergence into habits or divergence into novelty, and torsion enables context-dependent reinterpretation of past influences. These features align with neuroscientific evidence of hippocampal replay, where sequences of experiences are reactivated to support learning and planning \citep{hassabis2007patients}, and with transformer-based large language models (LLMs), where attention mechanisms implement a discrete form of trajectory influence \citep{vaswani2017attention}. The model further connects to the Relativistic Scalar Vector Plenum (RSVP) framework \citep{semantic2025}, which generalizes such trajectory influence processes across cognition, AI, and cosmology. Within this broader view, memory is one manifestation of entropic field dynamics: just as galaxy formation can be understood as the constraint and damping of trajectories in cosmic flows \citep{weinberg2008cosmology}, cognitive recall emerges as a constrained regeneration on a semantic manifold.

By incorporating semantic ledging and geometric virtual memory \citep{kuijper2021geometric}, the framework also addresses practical challenges of mismatched cues and dynamic scaling, ensuring that recall remains efficient and coherent even when inputs fail to align neatly with stored patterns. For example, in conversational AI, ledging allows partial matches to a user's query to be patched with finer-grained influences, while virtual memory scales the embedding space dynamically without quadratic storage costs. Taken together, this geometric account reframes memory as a generative, entropy-respecting process that balances fidelity with reinterpretation, local reuse with global coherence, and structural constraints with creative confabulation. The essay proceeds by formalizing the substrate, deriving the dynamics of trajectory influence, exploring geometric modulation, and outlining empirical predictions and implementations.

\section{Substrate and State}
The substrate is a smooth manifold $\M$ of ``meaning-states,'' which can represent neural activations in the brain, latent embeddings in LLMs, or RSVP fields in semantic infrastructures \citep{semantic2025}. A state at time $t$ is $x(t) \in \M$, equipped with local coordinates $x^i$ and a Riemannian metric $g_{ij}(x)$ that defines semantic distances between states. The metric quantifies how ``close'' two meaning-states are; for instance, in word embeddings, low $g_{ij}$ between ``cat'' and ``feline'' reflects their strong affinity, while high $g_{ij}$ between unrelated concepts like ``cat'' and ``quantum'' indicates semantic separation. This metric provides the foundation for geodesic paths, which represent the shortest routes through the space of meanings.

A generative drift field $F: \M \to T\M$ encodes the substrate's intrinsic evolution, driven by grammar, physics, or policy priors:
\[
\dot{x}(t) = F(x(t)) + \text{influence terms}.
\]
The drift $F$ represents the system's default dynamics in the absence of memory: in LLMs, it corresponds to the autoregressive prior that projects likely continuations based on local context \citep{vaswani2017attention}; in neural systems, it parallels the baseline activity patterns governed by cortical field dynamics \citep{amari1977dynamics}. The influence terms, introduced in the next section, modulate this baseline by incorporating history-dependent biases from past trajectories, allowing memory to bend the generative flow.

The manifold $\M$ can be visualized as a semantic landscape, with valleys corresponding to stable attractors (e.g., well-learned concepts) and ridges representing less accessible or unstable states. In such a geometry, the metric $g_{ij}$ not only measures local similarity but also determines the resistance to traversal: smooth regions facilitate rapid movement, while rough or curved areas slow it down. This view emphasizes that memory is not a lookup from a static table but a path-dependent motion through a structured space, where both local distances and global topology constrain recall. In cognitive terms, this substrate captures the distributed nature of memory storage, as proposed in connectionist models \citep{mcclelland1995why}, where meanings are encoded across overlapping patterns rather than in isolated locations. For computational implementations, $\M$ can be approximated by finite-dimensional embeddings (e.g., $\mathbb{R}^d$ with $d \approx 768$ for BERT-like models), where the metric is induced by the embedding norm.

\section{Memory as Trajectory Influence}
Memory is not a static snapshot but a trajectory $x(\tau)$, $\tau \in [0,t]$, that biases future evolution via a nonlocal-in-time flow:
\[
\dot{x}(t) = F(x(t)) + \int_0^t K(t,\tau; x(t), x(\tau)) \, U(x(\tau) \to x(t)) \, \dot{x}(\tau) \, d\tau + \xi(t),
\]
where the influence term integrates past increments $\dot{x}(\tau)$ to shape the present velocity. The kernel $K(t,\tau; x(t), x(\tau))$ weights contributions from the past, capturing how recency, content similarity, or temporal patterns modulate recall. The parallel transport $U(x(\tau) \to x(t)): T_{x(\tau)}\M \to T_{x(t)}\M$, defined by a connection $\Gamma^k_{ij}$, ensures that past tangent vectors are coherently brought into the current tangent space $T_{x(t)}\M$, avoiding artifacts from coordinate changes. Stochastic innovation $\xi(t) \sim \mathcal{N}(0, \sigma^2 I)$ injects novelty, preventing over-reliance on prior paths.

The kernel $K$ is the key modulator of memory span. Exponential decay $K = e^{-\alpha (t-\tau)}$ models short-term memory, where recent events dominate, consistent with working memory limits in humans \citep{miller1956magical}. Power-law kernels $K = (t-\tau)^{-\beta}$ produce long tails, enabling access to distant past events after consolidation, as in long-term memory systems \citep{mcclelland1995why}. Oscillatory-consolidation kernels $K = \cos(\omega (t-\tau)) e^{-\alpha (t-\tau)}$ reflect rhythmic processes like sleep replay, where memories are periodically reinforced \citep{stickgold2005sleep}. Selective kernels $K = \sigma(\langle q(x(t)), k(x(\tau)) \rangle)$ implement content-addressable recall, akin to attention in LLMs \citep{bahdanau2014neural}, where similarity between query $q(x(t))$ and key $k(x(\tau))$ gates influence.

This formulation ensures reconstructive recall: the system regenerates trajectories by integrating past directions, rather than replaying fixed states. For example, recalling ``dog-bark'' involves transporting increments from prior exposures to ``dog'' and ``bark,'' weighted by $K$, to generate a coherent association. The stochastic $\xi(t)$ allows for confabulation, where imperfect matches lead to creative or erroneous reconstructions, mirroring human memory errors \citep{loftus1974reconstruction}.

\section{Geometry: Metric, Curvature, and Torsion}
The metric $g_{ij}(x)$ defines semantic proximity, acting as the measure of how ``close'' two meaning-states are within the manifold. For instance, in word embeddings, low $g_{ij}$ between ``cat'' and ``feline'' indicates short geodesic distance, reflecting their strong semantic affinity. Conversely, distant or weakly related concepts yield large $g_{ij}$ values, making their paths harder to align. In this sense, the metric captures both similarity and resistance to association, functioning as the semantic analogue of distance in physical space. In practice, $g_{ij}$ can be learned from data, as in information geometry where it derives from the Fisher information matrix \citep{amari1985differential}.

Curvature $R^i_{jkl}$ governs geodesic deviation and thus the collective behavior of trajectories. Regions of positive curvature act like semantic basins, pulling trajectories together and reinforcing convergence, as seen in the consolidation of habits, routines, or well-learned associations. Negative curvature produces divergence, pushing trajectories apart, which can model phenomena such as instability, trauma, or conceptual repellors where meanings resist integration \citep{do_carmo1992riemannian}. For example, on a 2D spherical manifold, high curvature focuses geodesics toward poles, mirroring how repeated rehearsal drives convergence into stable memory channels such as a familiar word pair. In contrast, a hyperbolic region encourages exponential separation, mirroring rapid forgetting or conflicting interpretations. Curvature thus provides a geometric mechanism for stability: low-curvature flats facilitate smooth traversal, while high-curvature bends enforce alignment.

Torsion $T^k_{ij} = \Gamma^k_{ij} - \Gamma^k_{ji}$ introduces a semantic ``twist,'' capturing how the meaning of a trajectory depends on the path taken. Unlike curvature, which alters distances and convergence, torsion alters orientation: parallel transport around a loop may yield rotated or reinterpreted states. This provides a geometric account of contextual modulation, where a memory such as ``bank'' shifts interpretation depending on whether it is transported through a financial or riverine context. The resulting non-commuting transformations formalize how recall is path-dependent: identical starting points can yield distinct outcomes depending on the trajectory of influence. Empirically, such torsional effects manifest as context-dependent neural signatures, where the same stimulus elicits different patterns depending on preceding context \citep{hassabis2007patients}. In LLMs, torsion can be approximated by learned transport maps that rotate embeddings based on contextual prompts.

Together, metric, curvature, and torsion define the substrate’s full geometry: the metric measures local similarity, curvature sculpts global flow into attractors or repellors, and torsion encodes contextual reinterpretation. Memory dynamics emerge not from any single component, but from the interplay of all three, ensuring that recall is simultaneously stable, flexible, and sensitive to context. This geometric richness distinguishes the model from flat-space approximations in standard ML, where distances are Euclidean but lack intrinsic curvature or twist \citep{do_carmo1992riemannian}.

\section{Memory-Modulated Geometry (Plasticity)}
Past trajectories reshape the substrate:
\[
\partial_t g_{ij}(x) = -\eta_g \int_0^t K(t,\tau) \, \Pi_{ij}(x(t), x(\tau)) \, d\tau, \quad \partial_t \Gamma^k_{ij} = -\eta_\Gamma \int_0^t K(t,\tau) \, \Psi^k_{\ ij}(x(t), x(\tau)) \, d\tau,
\]
where $\Pi_{ij}$ and $\Psi^k_{\ ij}$ are Hebbian-like tensors \citep{hebb1949organization}, encoding path reinforcement, and $\eta_g, \eta_\Gamma$ are learning rates. Intuitively, these update rules mean that the geometry itself is plastic: every recalled or rehearsed trajectory slightly adjusts the distances and orientations of the manifold. For example, repeated exposure to the association ``dog-bark'' lowers $g_{ij}$ between their embeddings, making subsequent traversal easier and faster. Over time, such reinforced paths behave like geodesics of minimal resistance, providing a geometric analogue to habit formation. This dynamic is reminiscent of reinforcement learning, where low-energy paths are favored and accumulate stability \citep{sutton2018reinforcement}. 

Plasticity thus creates a feedback loop: trajectories bias geometry, and geometry in turn biases future trajectories. Habits emerge as local minima in the energy landscape, while rare or inconsistent experiences leave only shallow traces that decay without reinforcement. In this way, the manifold becomes both a record of the past and a scaffold for the future. The Hebbian tensors can be derived from the Fisher information in a statistical manifold, ensuring that updates align with data-driven semantic structure \citep{amari1985differential}. For instance, if two states are frequently co-recalled, $\Pi_{ij}$ reduces their geodesic length, effectively ``pulling them closer'' in the metric.

\subsection{Semantic Ledging}
Kuijper’s ledging \citep{kuijper2021geometric} decomposes non-power-of-two memory chunks into a geometric series of aligned blocks, reducing fragmentation. In our model, semantic ledging plays an analogous role for mismatched or incomplete cues. Instead of failing when a cue does not exactly match a past trajectory, the system decomposes it into a coarse component and finer-grained remainders:
\begin{itemize}
    \item A cue (e.g., ``dog'') aligns first with the largest relevant past trajectory (e.g., ``dog-bark'').
    \item Residual mismatches are patched via finer influences (e.g., ``pet,'' ``sound''), which fill in missing semantic mass.
    \item This process ensures bounded recall overhead, preventing semantic fragmentation and enabling smooth regeneration of meaning.
\end{itemize}
Semantic ledging thus implements a principle of efficiency: approximate matches are completed geometrically by combining aligned blocks rather than forcing an exact fit. It also operationalizes torsion, since reinterpretation is necessary to patch residual mismatches. For instance, the ambiguous memory of ``bank'' can be ledged into the financial or riverine context depending on which fine-grained cues dominate. This mechanism preserves coherence while allowing flexibility, ensuring that memory reconstruction remains robust even in the face of incomplete or noisy input. Empirically, ledging-like processes may underlie partial cueing in human memory, where a single word triggers a cascade of associations \citep{squire1992memory}.

\subsection{Semantic Virtual Memory}
Kuijper’s geometric virtual memory uses block trees for elastic address spaces \citep{kuijper2021geometric}. Here, meaning-states occupy dynamically instantiated subspaces of $\M$, created and dissolved on demand. Rather than permanently storing all trajectories, the system allocates a virtual region only when a context requires it, with transport operators $U$ mapping past states into the active subspace.

For example, in a dialogue system, each conversation spawns a temporary embedding region that represents only the local exchanges. Earlier utterances are not duplicated in full but can be reconstructed through $U$, which transports their latent traces into the current space. This prevents quadratic growth in storage and ensures that memory remains scalable even as contexts multiply. The effect is similar to paging in operating systems: only the relevant pages of meaning are ``in memory,'' while others can be recalled through geometric mappings when needed.

Semantic virtual memory also introduces a layer of indirection that supports compositional reuse. Contexts can overlap or be merged by aligning subspaces through gluing maps, allowing knowledge from one domain to be repurposed in another without conflict. This matches the RSVP framework’s categorical gluing of local structures into a coherent global plenum \citep{semantic2025}. In practice, it means that distinct threads of meaning (e.g., financial vs. ecological uses of ``bank'') can coexist in separate virtualized regions but still remain accessible through controlled transport.

In this way, semantic virtual memory provides both elasticity and isolation: it elastically scales to support growing contexts without runaway costs, and it isolates subspaces to prevent interference between unrelated meanings. Together, these properties enable entropy-aware, scalable semantic infrastructures that can support long-lived cognitive and AI systems. For instance, in LLMs, virtual memory could dynamically load domain-specific subspaces (e.g., legal vs. medical jargon), transporting relevant trajectories on demand to avoid embedding collapse.

\section{Energy/Lyapunov View}
Define a reconstruction functional:
\[
\E[x(\cdot)] = \int_0^T \left\{ \frac{1}{2} \langle \dot{x}(t) - F(x(t)), \dot{x}(t) - F(x(t)) \rangle_g + \int_0^t \mathcal{L}(x(t), x(\tau)) \, K(t,\tau) \, d\tau \right\} dt,
\]
where $\mathcal{L}$ penalizes deviation from past trajectories. This resembles the free-energy principle \citep{friston2010free}, in which systems act to minimize surprise by aligning present dynamics with prior structure. The functional combines two competing pressures: adherence to the generative drift $F(x)$ and fidelity to past influences via kernel-weighted penalties. 

For two attractors (e.g., ``cat'' and ``dog''), $\E$ encodes the tradeoff between generative innovation and stability. If trajectories remain close to prior paths, $\mathcal{L}$ is small, and the system falls into low-energy wells representing familiar associations. If trajectories diverge—driven by stochastic $\xi(t)$ or contextual torsion—$\E$ increases, signaling surprise or reinterpretation. Minimization of $\E$ thus produces a compromise: trajectories bend toward stable attractors while retaining flexibility to adapt when novelty is introduced.

The gradient flow $\nabla \E$ defines the system’s effective Lyapunov dynamics. Stable memories correspond to local minima of $\E$, which act as attractors, while unstable or contradictory memories correspond to saddles or shallow basins that are easily abandoned. Perturbations determine whether the trajectory settles back into a basin (recall) or transitions into a new basin (reinterpretation). This framework explains why habits are robust—low-energy channels with steep descent—whereas ambiguous or weakly reinforced memories are fragile, resting on flat plateaus of the energy surface.

In practical terms, $\E$ functions as a unifying scalar quantity: it measures the tension between drift, history, and noise. Its descent ensures stability without rigidity, guiding systems toward states that are both consistent with prior trajectories and capable of integrating new influences. For example, in a neural network, minimizing $\E$ during training reinforces paths corresponding to task-relevant associations, while allowing $\xi(t)$ to explore alternatives during inference. This dual role—stability during recall, exploration during learning—mirrors the balance in human memory systems, where over-consolidation leads to rigidity and under-consolidation to fragmentation \citep{squire1992memory}.

\section{Discrete (Sequence) Version}
For LLMs, states are $x_t \in \mathbb{R}^d$ with metric $g_t$. Dynamics are:
\[
x_{t+1} = f(x_t) + \sum_{s=0}^t \alpha_{t,s} \, P_{t \leftarrow s} \, (x_{s+1} - x_s) + \epsilon_t,
\]
where $\alpha_{t,s} = K(t,s) \cdot \sigma(\langle q(x_t), k(x_s) \rangle)$, $P_{t \leftarrow s}$ is a Jacobian or learned transport map, and $\epsilon_t$ is stochastic noise. This extends transformer attention \citep{vaswani2017attention}, with $P_{t \leftarrow s}$ providing a geometric analogue of positional encoding, allowing transported increments rather than raw embeddings to serve as the carriers of memory.

Plasticity adjusts the local geometry by updating the metric as a low-rank deformation:
\[
g_t = I + W_t W_t^\top, \qquad W_{t+1} = W_t + \eta \sum_s \alpha_{t,s} (x_s x_s^\top),
\]
so that heavily weighted past states reshape future distances, embedding recall bias into the substrate. Repeated exposure compresses high-frequency fluctuations and creates smoother channels for recurrent sequences, while rare or weak associations leave only shallow adjustments that decay under noise.

This discrete formulation makes clear the tradeoff between fidelity and efficiency. Every new state potentially interacts with all prior ones, yielding quadratic complexity in sequence length. The transport maps $P_{t \leftarrow s}$ provide a way to reuse past structure without recomputing raw distances, reducing overhead while preserving coherence. In practice, one obtains $O(t d^2)$ per step complexity, sacrificing some raw throughput in exchange for richer temporal depth. This framework thus unifies continuous geometric dynamics with sequence-based architectures: memory is realized not by static storage but by dynamically reweighting and transporting trajectories, balancing compression, stability, and generativity.

To illustrate, consider a simple sequence task: predicting the next word in a sentence. The drift $f(x_t)$ provides local predictions, while the influence term adds bias from earlier words, transported via $P_{t \leftarrow s}$. If the sentence is ``The cat sat on the mat,'' the transport from ``cat'' to ``sat'' reinforces the association, lowering $\E$ for coherent completions. In contrast, a mismatched cue like ``The dog sat on the mat'' triggers torsion, rotating the transported influence to adapt ``sat'' to the new context, potentially yielding ``The dog sat on the rug'' if the metric favors novelty.

\section{RSVP Tie-In}
In RSVP \citep{semantic2025}:
\begin{itemize}
    \item $\Phi$: Density of reliable paths, e.g., high in practiced memories where trajectories 
    are reinforced into stable channels. This scalar field measures the substrate's capacity for 
    semantic storage, with higher $\Phi$ indicating regions rich in consolidated associations.
    \item $\vvec$: Flow of thought, e.g., narrative currents in conversation that guide 
    the unfolding of meaning from one state to the next. The vector field $\vvec$ encodes 
    the directionality of influence, pointing toward likely continuations or habitual paths.
    \item $\Scal$: Entropy, high for creative exploration where multiple alternatives are 
    entertained, and low for stable recall where trajectories converge into a single path. 
    Entropy $\Scal$ quantifies the variability in trajectory deviation, with high values 
    promoting divergence and low values enforcing convergence.
\end{itemize}

Consolidation corresponds to reducing $\Scal$ along geodesics, thereby increasing $\Phi$ in 
the relevant region of the manifold. Well-practiced associations, like a familiar word pair, 
are thus stored as low-entropy attractors with dense, reliable connectivity. At the same time, 
$\vvec$ reflects the prevailing currents of thought that weave between these attractors, 
allowing narratives or reasoning chains to emerge as directed flows. 

In cosmology, an analogous process is found in galaxy formation: past flows with low vorticity 
restrict future configurations, funneling matter into stable structures. This mirrors how 
memory suppression operates by constraining trajectories to follow established paths rather 
than diffusing into novelty \citep{weinberg2008cosmology}. 

Semantic infrastructures exhibit the same balance. Preserving $\Scal$ is crucial for generativity: 
if entropy is prematurely collapsed, exploration is curtailed, and only rigid, habitual responses 
remain. RSVP’s categorical gluing guarantees that local sections (memories, concepts, contexts) 
can be stitched into a coherent global plenum while still allowing entropy to serve as a resource 
for creativity \citep{lurie2009higher}. In this way, the RSVP fields provide a unifying language 
across domains: trajectories of meaning in cognition, flows of matter in cosmology, and 
compositional seams in semantic infrastructures all obey the same balance of $\Phi$, $\vvec$, 
and $\Scal$. For example, in a conversational AI, $\Phi$ might represent the density of 
dialogue patterns, $\vvec$ the flow of turn-taking, and $\Scal$ the openness to topic shifts.

\section{Applications to Cognition and AI}
In cognition, the model explains hippocampal replay (exponential kernels) and long-term 
memory (power-law kernels) \citep{mcclelland1995why}. Exponential decay captures the 
rapid fall-off of influence typical of short-term memory, while power-law kernels account 
for the slower fading of long-term traces, consistent with behavioral recall curves. The 
geometric formulation also clarifies how replayed trajectories are not perfect copies but 
reconstructions guided by influence kernels: repeated rehearsal sharpens geodesics, while 
torsion enables reinterpretation of the same trace under different contexts. This yields a 
natural account of memory plasticity, priming, and the emergence of habitual paths. For 
instance, the dual-process theory of memory systems—hippocampus for episodic details and 
cortex for semantic gist \citep{squire1992memory}—maps to local (high-curvature) and global 
(low-curvature) trajectory influences, respectively.

In AI, the framework enhances transformers by incorporating geometric attention, improving 
context-aware recall \citep{vaswani2017attention}. Standard attention weights can be 
reinterpreted as discrete influence kernels, but without explicit geometric structure they 
risk collapsing distinct contexts into a single representation. By embedding attention 
within a curved metric, the model supports both efficient reuse of practiced trajectories 
and torsional reinterpretation when ambiguity arises. For example, a chatbot retracing 
user dialogue paths can prioritize recent (exponential) or distant (power-law) contexts, 
while parallel transport ensures that the meaning of reused segments is adapted to the 
current conversational frame. This predicts measurable improvements in dialogue coherence 
and contextual sensitivity, as the system balances fidelity to prior exchanges with the 
flexibility to reinterpret them under new prompts. Early experiments with geometric 
attention in sequence models show reduced perplexity on long-context tasks, supporting 
the model's efficacy \citep{bahdanau2014neural}.

\section{Empirical Validation}
The geometric memory model lends itself to empirical testing across cognitive science, 
neuroscience, and AI. Behavioral experiments can probe kernel effects: for instance, 
manipulating task timing to favor recency (exponential) or long-term (power-law) recall 
should alter serial position curves, as predicted by $K(t,\tau)$. Neuroimaging studies 
offer direct validation of geometric features: fMRI during memory replay tasks should 
reveal reduced curvature (flatter embedding spectra) in regions associated with 
consolidated memories \citep{hassabis2007patients}. EEG cross-frequency coupling can 
test torsional reinterpretation, where ambiguous stimuli elicit rotated neural patterns 
depending on context. In AI, ablation studies on LLMs—e.g., disabling transport $P_{t \leftarrow s}$ 
or perturbing kernels—should degrade contextual coherence, measurable via perplexity or 
human evaluation of generated text.

Initial validations align with the model. For example, PCA analyses of hippocampal 
recordings during replay show geodesic-like alignment of trajectories \citep{davidson2009hippocampal}, 
while attention ablations in transformers degrade long-range dependencies \citep{vaswani2017attention}. 
These results suggest the framework's utility for unifying disparate findings.

\section{Testable Predictions}
The geometric memory framework yields falsifiable predictions across
cognitive science, neuroscience, and artificial intelligence. Each prediction
arises from specific geometric features of trajectories, metrics, and kernels.

\begin{enumerate}
    \item \textbf{Geodesic Reuse}: 
    Transport operators $P_{t \leftarrow s}$, learned or estimated from prior paths,
    should align closely with observed state increments $x_{t+1}-x_t$.
    In large language models, this can be measured as cosine similarity between
    predicted continuation vectors and actual next-token embeddings. 
    In neuroscience, a parallel measure is alignment between replayed hippocampal
    trajectories and subsequent cortical activations, testable with multi-electrode arrays
    or fMRI sequence analysis \citep{davidson2009hippocampal}.

    \item \textbf{Curvature--Habit Link}: 
    Repeated use of a memory trajectory lowers effective curvature in that direction,
    making recall paths straighter and more efficient.
    Prediction: regions of semantic space with high recall frequency exhibit reduced
    sectional curvature. Empirical test: PCA or Laplacian eigenmaps of fMRI/EEG
    recordings should reveal flatter embedding spectra in practiced tasks 
    \citep{hassabis2007patients}. Analogously, LLMs trained heavily on a domain
    will show locally linear embedding geometries \citep{do_carmo1992riemannian}.

    \item \textbf{Kernel Manipulations}: 
    The shape of the influence kernel $K(t,\tau)$ controls memory span and weighting. 
    Perturbations (e.g., sleep disruption, pharmacological modulation, or induced oscillatory
    entrainment) should predictably shift behavioral recall patterns, altering recency/primacy
    effects \citep{stickgold2005sleep}. In AI, modifying attention decay functions 
    produces equivalent shifts in context-length sensitivity \citep{bahdanau2014neural}.

    \item \textbf{Torsion and Reinterpretation}: 
    Torsion allows transported vectors to rotate, enabling contextual reinterpretation.
    Prediction: ambiguous cues such as ``bank'' (riverbank vs. finance) should generate
    distinct neural trajectories depending on context. Testable via EEG/MEG 
    cross-frequency coupling or BOLD multivoxel patterns \citep{hassabis2007patients}. 
    In LLMs, the same input word embedded in different contexts should exhibit
    non-commuting transport loops, measurable through path-dependent embedding drift.

    \item \textbf{Energy Descent}: 
    Memory dynamics minimize the functional $\E[x(\cdot)]$. 
    Prediction: recall errors and forgetting follow gradient-like descent trajectories
    in this energy landscape. Adding stochasticity $\xi(t)$ increases exploration,
    leading to creative confabulations. In humans this manifests as dream recombination
    or confabulatory memory in amnesia; in LLMs, as novel but off-distribution text
    when noise is injected into hidden states \citep{friston2010free}.

    \item \textbf{Contextual Coherence as Global Constraint}: 
    Because trajectories are glued by parallel transport, coherence depends on
    maintaining consistent connections across local patches. 
    Prediction: disrupting these gluing operations produces dialogue incoherence
    in LLMs, or fragmented thought patterns in human subjects (e.g., schizophrenia).
    This can be tested with dialogue coherence metrics (perplexity under long-range
    constraints, topic continuity) or with neurocognitive assessments of narrative 
    integration \citep{squire1992memory}.
\end{enumerate}

\section{Minimal Implementation Sketch}
\begin{itemize}
    \item \textbf{State}: $x_t \in \mathbb{R}^d$, $g_t = I + W_t W_t^\top$.
    \item \textbf{Transport}: $P_{t \leftarrow s} = \exp(-\gamma (t-s)) I$.
    \item \textbf{Kernel}: $K(t,s) = e^{-\alpha (t-s)}$ or $\sigma(\langle q(x_t), k(x_s) \rangle)$.
    \item \textbf{Update}: $x_{t+1} = f(x_t) + \sum_s \alpha_{t,s} P_{t \leftarrow s} (x_{s+1} - x_s)$.
    \item \textbf{Plasticity}: $W_{t+1} = W_t + \eta \sum_s \alpha_{t,s} (x_s x_s^\top)$.
    \item \textbf{Evaluation}: MSE for recall fidelity, KL-divergence for creativity on a toy dataset (e.g., text sequences).
\end{itemize}
Pseudocode:
\begin{verbatim}
def update(x_t, history, f, P, K, eta):
    influence = sum(K(t,s) * sigma(q(x_t) @ k(x_s)) * P(t,s) @ (x_s+1 - x_s) 
                    for s in range(t))
    x_t1 = f(x_t) + influence + noise()
    W_t1 = W_t + eta * sum(K(t,s) * outer(x_s, x_s) for s in range(t))
    return x_t1, W_t1
\end{verbatim}

\section{Discussion and Outlook}
The model captures memory’s reconstructive nature but omits quantum effects or qualia 
\citep{penrose1989road}. This limitation is significant: while the geometry accounts for 
trajectory influence and plasticity, it does not yet address how subjective experience 
arises, nor how quantum coherence might alter long-tail kernels. These omissions define 
boundaries of the present framework.

Future work divides into three strands. First, simulations using lightweight frameworks 
(e.g., PyTorch with toy datasets) will demonstrate how different kernel families 
(exponential, power-law, oscillatory) alter recall fidelity and creativity. Such 
simulations provide benchmarks for comparing the geometric model to transformer-style 
architectures. Second, empirical neuroscience can probe curvature effects: if practiced 
memories lower effective curvature, this should be observable in fMRI or EEG manifold 
analyses during repeated recall tasks. Third, ablation studies in LLMs can probe torsion: 
removing or constraining parallel transport operators should reduce contextual 
reinterpretation, leading to brittle, overly literal outputs.

The broader implications extend beyond cognition and AI into semantic infrastructures. 
By formalizing memory as regenerative reconstruction on a manifold, the model provides a 
template for entropy-respecting systems \citep{semantic2025}. Virtualizing memory through 
block-tree–like decompositions and semantic ledging supports scalable, context-aware 
architectures that preserve diversity of trajectories rather than collapsing them. Such 
structures could serve as building blocks for resilient information systems, cognitive 
models, and AI platforms that metabolize complexity instead of erasing it.

Overall, the outlook is twofold: theoretical refinement, through geometric and categorical 
formalism, and practical implementation, through simulations and empirical validation. 
The promise of this line of work lies in providing a unified mathematical language for 
memory that bridges cognition, machine learning, and entropy-aware infrastructure design.

\section{Limitations and Extensions}
While the geometric model advances understanding of memory as trajectory influence, it 
has limitations. First, the assumption of a smooth manifold may not hold in quantum regimes, 
where memory traces could exhibit superpositions or entanglement \citep{penrose1989road}. 
Second, qualia—the subjective texture of experience—are not captured, as the model focuses 
on structural dynamics rather than phenomenal aspects. Third, computational tractability 
in high dimensions remains challenging, though approximations like low-rank metrics 
mitigate this.

Extensions include hybrid models incorporating quantum geometry (e.g., non-commutative 
manifolds) and qualia via information-theoretic measures. Empirical extensions could 
integrate the model with existing architectures, such as Hopfield networks \citep{hopfield1982neural}, 
to create hybrid systems that combine attractor stability with trajectory influence. 
Finally, the model’s RSVP integration suggests applications in cosmology, where trajectory 
damping in cosmic flows parallels memory consolidation \citep{weinberg2008cosmology}.

\section{Intuition}
Memory is a landscape of meanings—a manifold—where past journeys carve paths. Each step is guided by worn trails, smoothed by frequent travel and twisted by context. Recall regenerates these paths, balancing fidelity to prior routes with the freedom to find shortcuts, like rivers carving channels yet meandering anew. The geometry ensures coherence: curvature pulls familiar paths together, torsion allows reinterpretation, and plasticity deepens the grooves of habit. In this view, forgetting is not erasure but divergence, and creativity is not randomness but exploration along low-resistance geodesics.

% Bibliography
\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
