\documentclass[11pt]{article}
\usepackage{amsmath,amssymb,geometry}
\geometry{margin=1in}

\title{\textbf{Executive Summary:\\
Affine Quantum Deformation and the Geometry of Awareness}}
\author{Flyxion}
\date{}

\begin{document}
\maketitle

The Affine Quantum Deformation Principle (AQDP) identifies a fundamental feature
of quantum spacetime: the averaged affine connection differs from the
Levi--Civita connection of the averaged metric. This difference is encoded in
the Quantum Affine Shift Tensor $\mathcal{A}$, derived from the covariance of
the metric operator. The resulting deformed Einstein equations, geodesic flow,
and Raychaudhuri evolution describe how quantum uncertainty reshapes the causal
structure of spacetime.

The Relativistic Scalar--Vector Plenum (RSVP) theory extends this geometric
framework to cognition. Semantic content is modeled by $\Phi$, cognitive flow by
$\mathbf{v}$, and uncertainty by $S$, together inducing a semantic metric whose
geometry is deformed under uncertainty. Awareness is formulated as the
preservation of both metric and spectral invariants under the RSVP flow. This
definition is mathematically equivalent to Markov boundary maintenance, a key
construct in generative modeling and neuroscience.

A single variational principle produces both AQDP and RSVP dynamics, revealing a
shared origin for geometric deformation and semantic coherence. Coherence and
awareness arise as invariants of a system navigating intrinsic uncertainty,
whether physical or cognitive.

\end{document}
