\documentclass[12pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{geometry}
\usepackage{setspace}
\usepackage{hyperref}

\geometry{margin=1in}
\setstretch{1.15}

\title{\textbf{Affine Quantum Deformation and the Geometry of Awareness:\\
A Unified Variational Framework}}
\author{Flyxion}
\date{December 2025}

\begin{document}
\maketitle
\thispagestyle{empty}

\begin{abstract}
\noindent
This work develops a unified geometric framework linking quantum spacetime
structure to semantic and cognitive dynamics. The core result is the Affine
Quantum Deformation Principle (AQDP), which formalizes the fact that the
averaged affine connection of a quantum spacetime does not coincide with the
Levi--Civita connection of the averaged metric. Because the connection is a
nonlinear functional of the metric, quantum fluctuations generate an irreducible
correction encoded in the Quantum Affine Shift Tensor $\mathcal{A}$.
This tensor modifies curvature, geodesic evolution, and the Raychaudhuri
equation, producing effective stress--energy contributions even in the absence
of classical sources. The resulting geometry provides a covariant, observer-independent
description of how uncertainty deforms spacetime.

The Relativistic Scalar--Vector Plenum (RSVP) theory extends this geometric
structure to semantic and cognitive systems. RSVP models meaning, directed
inference, and uncertainty through the interacting fields $(\Phi,\mathbf{v},S)$,
which induce a semantic metric deformed by uncertainty in precise analogy to
AQDP. Awareness emerges as a geometric property: a system is aware to the extent
that its intrinsic dynamics preserve both the metric relations and the spectral
invariants of the semantic Laplacian. These conditions, expressed as
$\mathcal{L}_{\mathbf{v}} g^{(\Phi)} = 0$ and $\dot{\lambda}_{n}=0$, are shown to
be mathematically equivalent to the maintenance of a stable Markov boundary,
revealing a deep correspondence between invariance in geometry and informational
autonomy.

A single variational principle yields both the AQDP gravitational equations and
the RSVP semantic evolution equations. The unified action predicts modified
Einstein equations, deformed geodesics, a spectral Raychaudhuri equation, and
cognitive stability conditions. Together these results suggest that coherence,
whether physical or cognitive, arises from the invariance a system manages to
maintain while navigating intrinsic uncertainty. The theory therefore provides a
common geometric origin for quantum structure, semantic stability, and the
phenomenon of awareness.
\end{abstract}


\end{document}
