\appendix

\chapter{Philological Notes (Hebrew, Greek, Arabic, Cross-Tradition)}
This appendix collects philological and comparative notes that support the scriptural exegesis in the main chapters. 

\section*{Hebrew: Job 31:33}
The Hebrew verb \emph{kasah} (כָּסָה, “to cover, conceal”) is central to Job 31:33. 
Its semantic field includes both the concrete (covering physical matter such as waste) and the abstract (covering guilt, words, or meaning). 
The ambiguity allowed scribes to encode dual layers: a hygienic injunction and a hermeneutic strategy of concealment.  
Cognates in Aramaic and Ugaritic preserve similar double valences, reinforcing the idea that concealment carried ritual, moral, and epistemic weight.  
Robert Alter’s rendering emphasizes this interpretive density \citep{alter2010wisdom}.  
Clines’s commentary further explores how concealment in Job reflects broader Ancient Near Eastern practices of scarcity in revelation \citep{clines1989job}.  

\section*{Greek: The Parable of the Sower}
The Synoptic Gospels (Matthew 13, Mark 4, Luke 8) present the Parable of the Sower.  
Greek terms such as \emph{speirō} (“to sow”) and \emph{ge} (“earth, ground”) highlight the agricultural imagery, 
while the Stoic notion of the \emph{logos spermatikos} (“seminal reason”) infuses the parable with philosophical overtones.  
Seeds are replicators, soils are environments: a natural selection allegory that anticipates Darwinian logic and Bayesian inference, 
where each seed is a hypothesis tested by conditions of growth.  

\section*{Arabic: \emph{Iqraʾ}}
The Qur’anic command \emph{iqraʾ} (\emph{أقرأ}) admits seven attested meanings:  
\begin{enumerate}
  \item to read or study,  
  \item to rhyme or versify,  
  \item to draw near or return,  
  \item to be delayed or held back,  
  \item to become pregnant (retaining seed),  
  \item to menstruate (retaining blood),  
  \item to rise and warn.  
\end{enumerate}
These meanings collectively form a proto-algorithm of cognition: gathering, ordering, delaying, retaining, and broadcasting.  
They align with later computational metaphors such as buffering, queuing, and triggering.  

\section*{Cross-Tradition Parallels}
\begin{itemize}
  \item In Buddhism, the doctrine of \emph{pratītyasamutpāda} (dependent origination) resonates with the parable’s conditional growth: nothing thrives outside relational soil.  
  \item In Hinduism, the Vedic \emph{ṛta} signifies cosmic order restored through ritual, comparable to the algebraic balance of Al-Khwarizmi.  
  \item Indigenous oral traditions often employ concealment and delayed disclosure as pedagogical devices: wisdom is revealed only in cycles, preserving scarcity.  
\end{itemize}

---

\chapter{Full Equations and Proofs (PCT, Active Inference, CTM, RSVP)}
This appendix contains detailed derivations of the main formal systems referenced in the text.

\section*{Perceptual Control Theory}
Behavior is modeled as perception control:
\[
u(t) = G \big( r(t) - p(t) \big),
\]
where $r(t)$ is a reference signal, $p(t)$ the perceived signal, and $G$ the controller gain.  
Closed-loop stability analysis demonstrates that control requires minimizing the error $e(t) = r(t) - p(t)$ across environments.  

\section*{Active Inference}
Active inference minimizes variational free energy:
\[
F(q) = \mathrm{KL}(q(z)\,\|\,p(z|x)) - \log p(x).
\]
Proof: agents act to minimize surprise ($-\log p(x)$), approximating hidden causes $z$ with beliefs $q(z)$.  
This unifies Bayesian inference with embodied action.  

\section*{Conscious Turing Machine}
In the CTM, the probability of a chunk $c$ winning access to short-term memory is
\[
P(c \text{ wins}) = \frac{w_c}{\sum_i w_i},
\]
independent of processor location.  
Two axioms govern the system: the Global Broadcast Axiom (attention) and the Inspection Axiom (awareness).  

\section*{RSVP PDEs}
The RSVP dynamics are:
\[
\partial_t \Phi + \nabla \cdot (\Phi \mathbf{v}) = -\alpha S,
\]
\[
\partial_t \mathbf{v} + (\mathbf{v}\cdot\nabla)\mathbf{v} = -\nabla \Phi + \beta \nabla S,
\]
\[
\partial_t S + \nabla\cdot(S\mathbf{v}) = \gamma \Delta \Phi - \delta S^2.
\]
Fourier analysis reveals smoothing effects, with $-\delta S^2$ ensuring entropy decay and preventing torsion runaway.  

---

\chapter{Computational Complexity and Information Bounds}
\section*{Complexity Classes}
RSVP concealment thresholds can be viewed as NP-like barriers: below a given entropy bound, local reconstruction of global state is computationally infeasible.  

\section*{Information-Theoretic Bounds}
Define the entropy-dependent threshold $\theta(S)$.  
Any observable $O$ with $\|O\| < \theta(S)$ becomes irretrievable, amounting to lossy compression.  
This bound ensures structural scarcity of interpretation across observers.  

---

\chapter{Catalogue Entry — Coran Cubic Orthohedron Rectangular and Neat}
\section*{Description}
A physical artifact: a 3×2×19 bookshelf designed to house daily and evening scrolls across a Metonic cycle.  
The acronym “CORAN” encodes Cubic, Orthohedron, Rectangular, and Neat.  

\section*{Symbolism}
The geometry encodes concealment, restoration, and balance.  
Its form alludes simultaneously to scriptural cycles, algebraic symmetry, and cosmological recurrence.  

---

\chapter{The Seven Meanings of \emph{Iqraʾ} Poem}
\section*{Text}
\begin{quote}
Read and study, collect, piece it together and know.\\
Match portion to portion.\\
Draw near or go back after being away.\\
Cite, recite, quote.\\
Be behind, held back, or delayed.\\
Wait a month.\\
Get up and Warn.
\end{quote}

\section*{Analysis}
Each verse encodes one of the philological senses of \emph{iqraʾ}.  
Together they form a poetic algorithm of concealment, delay, and revelation.  
Parallels include Buddhist chanting, Vedic recitation, and liturgical cycles in Judaism and Christianity.  

---

\chapter{LaTeX Macros and Code for Simulations}
\section*{Macros}
\begin{verbatim}
\newcommand{\RSVPphi}{\Phi}
\newcommand{\RSVPv}{\mathbf{v}}
\newcommand{\RSVPs}{S}
\end{verbatim}

\section*{Python Simulation Snippet}
\begin{verbatim}
for t in range(T):
    Phi = Phi - alpha*S*dt - div(Phi*v)*dt
    v   = v - grad(Phi)*dt + beta*grad(S)*dt
    S   = S + gamma*laplacian(Phi)*dt - delta*S**2*dt
\end{verbatim}
Toy lattice models capture entropy smoothing and vector torsion suppression.  

---

\chapter{Extended Mathematical Infrastructure}
\section*{Category Theory}
RSVP formalized as a symmetric monoidal category: objects are field configurations, morphisms are entropy-respecting transformations.  

\section*{Sheaf Theory}
Sheaf-theoretic gluing encodes how local sections of reality must be combined to reconstruct the plenum.  
Entropy on overlaps prevents trivial gluing, enforcing viewpoint diversity.  

\section*{Topological Analysis}
Homotopy colimits guarantee coherence of global states.  
Spectral gaps in Fourier space reveal crystalline reset and Poincaré recurrence.  

---

\chapter{Bibliographic Annotations}
\begin{itemize}
    \item Alter: literary translation of Job as philological grounding.  
    \item Clines: commentary on Job’s concealment motifs.  
    \item Baars: Global Workspace Theory.  
    \item Blum \& Blum: Conscious Turing Machine.  
    \item Friston: Free Energy Principle.  
    \item Smolin: Cosmological Natural Selection.  
    \item Al-Khwarizmi: Algebraic restoration and balancing.  
    \item Mac Lane and Awodey: category and sheaf theory foundations.  
\end{itemize}
